Paper 4: Praise the people or praise the place?

### 2024-01-11 index.qmd
add write up? " git push

# Create the ggplot visualization
ggplot(books, aes(x = records, y = name, fill = factor(volume))) +
    geom_col(alpha = .8) +
    labs(
        x = "Number of biographies", y = "Book volume",
        title = "Number of biographies in Vem är Vem?",
        fill = "Edition"
    ) +
    scale_x_continuous(labels = scales::number_format()) +
    scale_fill_manual(values = c("#2B8CBE", "#be5d2b", "midnightblue"))

>>>>

# Create the ggplot visualization
ggplot(books, aes(x = records, y = name, fill = factor(volume))) +
    geom_col(alpha = .8) +
    labs(
        x = "Number of biographies", y = "Book volume",
        title = "Number of biographies in Vem är Vem?",
        fill = "Edition"
    ) +
    scale_x_continuous(labels = scales::number_format()) +
    scale_fill_manual(values = c("#2B8CBE", "#be5d2b", "midnightblue")) +
    theme(plot.title.position = "plot")

# ggsave("assets/who-is-who-volumes.png", width = 20, height = 20, units = "cm")

### 2024-01-11 paper.qmd
add write up? " git push

---
title: "Paper 4: Praise the people or praise the place"
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default

bibliography: bibliography.bib
---

@@

In the quest to understand the dynamics of economic development and technological advancement, previous research by this author and his supervisors shed light on the transformative impact of early electricity access in Sweden. "Power for progress: The impact of electricity on individual labor market outcomes" [@jayes_power_2024] revealed how the advent of electricity in certain parishes led to positive economic outcomes: a boost in income levels, reduced inequality, and the maintenance of employment levels despite the advent of labor-saving technology. A particularly striking observation was the tendency of workers in these electrified parishes to remain in their birthplaces, hinting at a newfound economic vibrancy stemming from the technological spillovers.

@@

The historical adaptability of labor markets to technological change is well-documented. In their study of the U.S. labor market's response to the automation of telephone operation, Feigenbaum and Gross [@Feigenbaum2020] demonstrate how technological displacement in one sector led to increased demand in others, suggesting an inherent resilience in labor markets. This finding is particularly pertinent to our exploration of Sweden's electrification, as it indicates a potential for both displacement and opportunity in the face of technological change.

Goldin's extensive analysis of labor markets in the 20th century provides a comprehensive backdrop to our study [@Goldin1994; @Goldin1998]. Her work highlights critical shifts in labor participation, wage structures, and job security, reflecting the complex interplay between societal changes and labor market dynamics. These insights are crucial for understanding how shifts in human capital, like those during Sweden's electrification period, contribute to broader economic outcomes.

@@

## Biographical dictionaries

"Vem är Vem?" is a biographical dictionary, comprising a rich repository of information about notable individuals in Sweden. Published in two editions with a total of five volumes each, the first edition spanned from 1945 to 1950, and the second from 1962 to 1968, by the Bokförlaget Vem är Vem publishing house. An additional volume specifically focussed on individuals in industry and business was produced in 1945. This encyclopedia offers an invaluable snapshot of Swedish societal and professional landscapes during these pivotal periods. [fix citation]

@@

The depth and breadth of "Vem är Vem?" are further enhanced by its digitization, with 9 out of the 11 volumes being made accessible online by librarians in Uppsala. This digitization has significantly facilitated research, allowing for a broader and more efficient exploration of the biographies and career paths of thousands of individuals. The encyclopedia's extensive coverage makes it a goldmine for researchers, historians, and anyone interested in the socio-economic history of Sweden during a period marked by significant change and development.

@@

![A representative page of Vem är Vem?](assets/vem_page.png)

## Industrial Catalogue

We make use of the Swedish Industrial Calendar (*Svensk Industrikalender*) from 1947. This document is digitized and made available online by *Projekt Runeberg*.

[finish description]

![](assets/svensk-industrikalender_1.png)

@@

1. **Company Name**: The name of the company is listed at the beginning of each entry, usually in bold or with an asterisk.

@@

9. **Address**: The full postal address, which may include a street name or a postbox number, indicated as "Postgiro".

This type of catalogue was commonly used for business-to-business interactions and could be considered an early form of networking resource, allowing companies to find suppliers, customers, and partners.

@@

We need to digitize this record in order to use it.

We make use

```{mermaid}
graph TD
    A[Start] --> B[Download Images]
    B --> C[Send Images to Azure OCR]
    C --> D[Receive JSON Files]
    D --> E[Analyze JSON for Geographic Clusters]
    E --> F[Identify Technology Clusters]
    F --> G[End]

@@

### Information about the company register:

[THE COMPANY REGISTER](https://runeberg.org/svindkal/1947/0140.html)

contains information regarding industrial companies' place of address, nature, main
productions, location of various factory departments, head office and
branch office postal and telegram addresses, telephone and postal giro, representatives in
important locations within the country; share capital for companies operating under
limited liability company form; approximate number of industrial workers or, where production is
temporarily significantly curtailed due to prevailing conditions, the workforce
normally employed; approximate manufacturing value,
manufacturing quantity or manufacturing capacity, where information about this could be
published; year of construction; the board members of the limited companies (in alphabetical order)
and the owners of other companies; names of managers and department heads to the
extent resp. company wished to let information about this influence.

A * denotes companies affiliated to Sweden's Confederation of Industry. For companies
that have an ad in the calendar's ad section, there is a reference to
the number or location of the ad.

25 people who are hisco code 23. We can look into where they studied and worked.

This is the guy we tell our story around:

 "3": {
            "full_name": "Nordell, Axel Verner",
            "location": "Motala, Östergötland",
            "occupation": "Civilingenjör, kraftverksdirektör",
            "birth_details": {
                "date": "15/08/1881",
                "place": "S. Möckleby, Kalmar",
                "parents": "Gustaf N. and Amanda Seillergren",
                "place_coordinates": {
                    "Latitude": 56.35646300000001,
                    "Longitude": 16.420155
                }
            },
            "education": [
                {
                    "degree": "Studentexamen",
                    "year": "1899",
                    "institution": "Lunds högre allmänna läroverk",
                    "institution_coordinates": {
                        "Latitude": 55.7046601,
                        "Longitude": 13.1910073
                    }
                },
                {
                    "degree": "Avgångsexamen från KTH",
                    "year": "1904",
                    "institution": "KTH (E)",
                    "institution_coordinates": {
                        "Latitude": 60.12816100000001,
                        "Longitude": 18.643501
                    }
                }
            ],
            "career": [
                {
                    "position": "Ritare",
                    "years": {
                        "start_year": 1904,
                        "end_year": 1905
                    },
                    "organization": "ASEA i Malmö"
                },
                {
                    "position": "Ingenjör",
                    "years": {
                        "start_year": 1905,
                        "end_year": 1907
                    },
                    "organization": "Elektr. A-B Hodmia i Sthilm"
                },
                {
                    "position": "Ingenjör",
                    "years": {
                        "start_year": 1907,
                        "end_year": 1909
                    },
                    "organization": "Trollhätte (kanal- och vattenverk)"
                },
                {
                    "position": "Chef för vattenfallsverket i Älvkarleby, sektion Motala",
                    "years": {
                        "start_year": 1918,
                        "end_year": 1920
                    },
                    "organization": "Älvkarleby kraftverk"
                },
                {
                    "position": "Kraftverksdirektör",
                    "years": {
                        "start_year": 1920,
                        "end_year": 1947
                    },
                    "organization": "Motala kraftverk"
                },
                {
                    "position": "Verkställande direktör",
                    "years": {
                        "start_year": 1930,
                        "end_year": 1947
                    },
                    "organization": "Motala Ströms Kraft A-B"
                }
            ],
            "family": {
                "spouse": "Agnies Hellgren",
                "marriage_year": "1911",
                "children": [
                    {
                        "name": "Inga",
                        "birth_year": "1912"
                    },
                    {
                        "name": "Hans",
                        "birth_year": "1914"
                    },
                    {
                        "name": "Gösta",
                        "birth_year": "1918"
                    },
                    {
                        "name": "Ulla",
                        "birth_year": "1920"
                    }
                ]
            },
            "location_coordinates": {
                "Latitude": 58.5380335,
                "Longitude": 15.0470936
            },
            "occupation_hisco_code": 34,
            "occupation_cosine_similarity": 0.7726821043535681,
            "occupation_hisco_occupation": "Electrical and Electronics Engineering Technicians"
        }
    }
}

>>>>

---
title: "Paper 4: Praise the people or praise the place?"
subtitle: Upper tail human capital in electrifying Sweden
code-repo: "Access the code, data, and analysis at <https://github.com/j-jayes/who-is-who-scraper> and <https://github.com/j-jayes/svensk-industrikalender>"
author:
  - name: Jonathan Jayes
    email: jonathan.jayes@ekh.lu.se
    orcid: 0000-0003-4967-4869
    affiliations: 
      - id: lu
        name: Lund University Economic History Department
abstract: |
  Abstract
date: January, 2024
thanks: |
  Thank you to seminar participants at the Copenhagen Business School Department of Strategy and Innovation PhD Seminar  and the HEDG Group at the University of Southern Denmark for valuable feedback on this work.
format:
  hikmah-pdf: default
  hikmah-manuscript-pdf: default
execute: 
  echo: false
bibliography: bibliography.bib
---

@@

In the quest to understand the dynamics of economic development and technological advancement, previous research by this author and his supervisors shed light on the transformative impact of early electricity access in Sweden. "Power for progress: The impact of electricity on individual labor market outcomes" [@jayes_power_2024] revealed how the advent of electricity in certain parishes led to positive economic outcomes: a boost in income levels, reduced inequality, and the maintenance of employment levels despite the advent of labor-saving technology. A particularly striking observation was the tendency of workers in these early electrified parishes to remain in their birthplaces, hinting at a newfound economic vibrancy stemming from the technological spillovers.

@@

The historical adaptability of labor markets to technological change is well-documented. In their study of the U.S. labor market's response to the automation of telephone operation, Feigenbaum and Gross @Feigenbaum2020 demonstrate how technological displacement in one sector led to increased demand in others, suggesting an inherent resilience in labor markets. This finding is particularly pertinent to our exploration of Sweden's electrification, as it indicates a potential for both displacement and opportunity in the face of technological change.

Goldin's extensive analysis of labor markets in the 20th century provides a comprehensive backdrop to our study [-@Goldin1994; -@Goldin1998]. Her work highlights critical shifts in labor participation, wage structures, and job security, reflecting the complex interplay between societal changes and labor market dynamics. These insights are crucial for understanding how shifts in human capital, like those during Sweden's electrification period, contribute to broader economic outcomes.

@@

## Source material

### Biographical dictionaries

"Vem är Vem?" is a biographical dictionary, comprising a rich repository of information about notable individuals in Sweden. Published in two regional editions with a total of five volumes each, the first edition spanned from 1945 to 1950, and the second from 1962 to 1968, by the Bokförlaget Vem är Vem publishing house. An additional volume specifically focussed on individuals in industry and business was produced in 1945. This encyclopedia offers an invaluable snapshot of Swedish societal and professional landscapes during these pivotal periods. [fix citation]

@@

It is worth noting that the criteria for inclusion was somewhat vague, and individuals could opt in to being included for a nominal fee. As a result, there are some individuals for whom not much information is included beyond biographic information, current location and profession. For others, there is a rich tapestry about their lives including records of career progression, business travel, technical writings and membership of civic organizations. The source does not capture a representative picture of Swedish society at the time, but rather those individuals with some level of social cachet or prestige, and a desire to be recorded in the biographical dictionaries as such.

"Vem är Vem?" is useful to economic historians thanks to its high quality digitization, with nine out of the 11 total volumes being made accessible online by librarians in Uppsala through *Projekt Runeberg*, as shown in @fig-who-is-who-books. This digitization has facilitated research, allowing for a broader exploration of the biographies and career paths of thousands of individuals. The encyclopedia's extensive coverage makes it a goldmine for researchers, historians, and anyone interested in the socio-economic history of Sweden during a period marked by significant change and development.

```{r}
#| label: fig-who-is-who-books
#| fig-cap: "Number of biographies in each volume of 'Vem är Vem?'"
#| fig-width: 5
#| out-width: 5in

knitr::include_graphics("assets/who-is-who-volumes.png")
```

@@

The biographic information about the individuals in the dictionaries are exemplified in @fig-who-is-who-page, which highlights the life of chemist and metallurgist Karl Gustaf Lund.

```{r}
#| label: fig-who-is-who-page
#| fig-cap: "A representative page of Vem är Vem?, highlighting the biography of Karl Gustaf Lund"

knitr::include_graphics("assets/vem_page.png")
```

The fields include:

1.   **Education**: Lund’s education at prestigious institutions such as the Royal Institute of Technology (KTH) indicates he had access to advanced technical knowledge. This level of education is critical for understanding the specialized skills that were necessary for innovation and advancement in electricity-related industries.

2.   **Career Progression**: The text outlines Lund’s career progression through various roles in metallurgy and chemical engineering. This trajectory can illustrate how individuals applied their education in practice, contributing to industrial development. Tracking such careers can provide insight into the professional development paths that were common and valued in the sector at the time.

3.   **International Experience**: His experiences in the United States reflect the cross-border exchange of knowledge and skills. This can show how international experiences contributed to the domestic industry by importing new ideas and practices, which is a key aspect of human capital development.

4.   **Leadership and Management**: Lund’s leadership positions, such as chairmanships and advisory roles, imply a combination of technical expertise and managerial acumen. The ability to lead and innovate within companies is a significant aspect of human capital that drives industry growth.

5.  **Research and Innovation**: The reference to his translated research work indicates an engagement with cutting-edge technology and knowledge creation. Such contributions are the tangible outputs of human capital in action, pushing the industry forward through innovation.

6.   **Professional Networks**: His involvement with societies and associations suggests a networked professional community, which is essential for the diffusion of innovative ideas and practices. These networks are often where knowledge is exchanged, partnerships are formed, and collaborations are initiated.

### Industrial Catalogue

The Svensk Industrikalender from 1947, published by Sveriges Industriförbund (Sweden's Industrial Association), is a comprehensive directory of Swedish industrial firms. This calendar was issued annually from 1918 to 2000 and contains information related to Swedish industry. The 1947 edition available on the Project Runeberg website was digitized in April 2012, sourced from the Centrum för Näringslivshistoria. The calendar is believed to be under catalog protection but not copyright [citation].

It includes detailed information such as company names, locations, nature of businesses, products, contact details, share capital, number of employees, production values, establishment years, and key personnel including managing directors and board members. This source is valuable for studying the economic and industrial environment of post-war Sweden, providing insights into corporate structures, industry distribution, and business trends of that period.

A representative page is shown in @fig-industry-calender.

<br>

```{r}
#| label: fig-industry-calender
#| fig-cap: "A page of the industry catalogue from 1947"
#| fig-width: 6
#| out-width: 6in

knitr::include_graphics("assets/svensk-industrikalender_1.png")
```

@@

1. **Company Name**: The name of the company is listed at the beginning of each entry, with an asterisk indicating membership of Sveriges Industriförbundet.

@@

9. **Address**: The full postal address, which may include a street name or a postbox number, indicated as `Postgiro`.

This type of catalogue was commonly used for business-to-business interactions and can be considered an early form of networking resource, allowing companies to find suppliers, customers, and partners.

@@

In order to analyze both the biographical dictionaries and industrial catalogue, we need to bend the text into a machine readable structure. This process is relatively complicated. It involves breaking each component of the source up (e.g. each biography or company record), extracting the pertinent information from each record, storing each value with its associated key, and then saving this information in a way that is easy to analyze and aggregate. The simplified process is laid out in @fig-data-collection-process.^[The underlying code can be found on the GitHub repo linked at on the first page of this paper.]

```{r}
#| label: fig-data-collection-process
#| fig-cap: "Data collection process steps"
#| fig-width: 5
#| out-width: 5in

knitr::include_graphics("assets/data-collection-process.png")

@@

Prior to the advent of Large Language Models (LLMs), this was a task that required a large number of human hours to complete, either putting the information into an excel sheet by hand, or writing rules to extract the information from the text. The first approach limits the number of observations a researcher can collect on her own, and the second approach quickly turns into the first.

Due to the number of abbreviations, acronyms and contractions (for example, `Gävleb. l.` is the contraction of *Gävleborg län* or Gävleborg county in @fig-who-is-who-page), while it might be possible to take a simple rules based approach to replacing these contractions with their complete Swedish text, and then looking for specific terms relating to each piece of information, the number of rules soon balloons to an unreasonable figure. Writing a rule for every case necessitates as much human involvement as would be required to manually structure the information - the first approach.

However, with the rapid advancements in LLM technology in the previous five years, and poplular adoption of these tools through Chat-GPT and Microsoft's integration of GPTs into their products in the previous year, new tools mean this manual workload can be avoided to a large extent.

I make use of the backed of Chat-GPT, a model called GPT-3.5-turbo from OpenAI to structure the information from the dictionaries and catalogue into a JSON format that I can analyse.

### Intuitive explanation of LLMs contextual 'understanding'

The GPT-3.5-turbo model which I make use of is a LLM which has been trained on all of Wikipedia and Wikidata, among others. These two sources contain the same information, but in a different format, as shown in the adapted extracts below.

<blockquote>

Jonas Wenström (4 August 1855 in Hällefors – 22 December 1893 in Västerås) was a Swedish engineer and inventor, who in 1890 received a Swedish patent on the same three-phase system independently developed by Mikhail Dolivo-Dobrovolsky. He studied at Uppsala University.

</blockquote>

```{r}
library(tibble)

wenstrom_data <- tibble(
  key = c("Name", "Birth Date", "Death Date", "Occupations", "Education"),
  value = c("Jonas Wenström", "1855-10-04", "1893-12-21", "Engineer, Inventor", "Uppsala University")
)

library(gt)

wenstrom_data %>% 
gt() %>% 
cols_label(key = "Key", value = "Value") %>% 
tab_header(title = md("Wikidata information about Swedish inventor **Jonas Wenström**"))

<br>

## Clustering of firms and classification of occupations

[explain how I cluster firms with text embeddings based on business and products]

[explain how I ]

## What does the industrial calendar tell us about industrial firms in Sweden in 1947?

![](assets/geom_density_ridges.png)

![](assets/map_most_common_businesses.png)

![](assets/map_of_electricity_related_businesses.png)

![](assets/capital_vs_employees.png)

## What can we learn from the Who is Who? biographies?

![](assets/hisco_counts.png)

![](assets/eee_map.png)

![](assets/distances_comparison.png)
